Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-07-24 analysis/paper/paper.Rmd
editing new text on size

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics are widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with clear distinct components [@Selden2019; @Topi2017]. With univariate or multivariate statistics, the GMM applications can answer research questions related to lithic topological evolution and variations [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019]. A common procedure is using dimensional reduction techniques, such as principal components analysis or canonical variate analysis, to capture the essential features that can represent the overall shape. This enables the identification of clustered analytical groups through the visualization of results, followed by statistical tests to robustly distinguish them.

One of the interesting topics is to explore whether and why there is shape standardization over time, or between assemblages across geographical areas. It is common to use multivariate analysis on shape variables or coefficient of variations on associated metric data to discuss standardization, especially for lithic assemblages. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to a maintenance of finished points. With considerations of a set of metric data, @Buchanan2018; @Buchanan2019 identified a more uniform base-shape of Folsom points than Clovis points across the western US, which indicates the specific function for hafting. Similarly, @Smith2016 suggested that the humongous fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel. Other factor, such as low level of cultural innovation in a small group could also lead to an increase in standardization of point shapes [@Okumura2014]. For ceramics, @Topi2017 found a standardized vessel shape from the Casas Grandes culture in northwest Mexico using coefficient of variations for the positions of semi-landmarks across shape groups. They suggest the standardization might indicate the pretense of specialized producers reflecting social complexity. Another way to examine standardization is using morphological disparity to perform pairwise comparison of correlations between shape groups [@Wills2001]. In this manner, @Selden2018ceramic2; @Selden2019 examined Caddo ceramics in northeast Texas region and found a gradual shape standardization over time that provides a basis for further discussion of craft specialization or group identity. Similar applications to bifaces in central Texas presented the evidence of grater diversity in morphology [@Selden2020]. More other applications, such as cranial deformation, demonstrate that landmark approach with multivariate analyses on shape variances or symmetry is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].

Despite the wide use of landmark approach in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, outline approach, such as Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in forms from a more angular to rounded shape over time. Later, @Saragusti2005 introduced several functions for the further calculation of the specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential for analysis of curves. To compare resharpening trajectories in the European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization to EFA, with considerations of size using multivariate linear regression. More recent applications to stone artifacts seek to answer questions about topological classification of Late Woodland points [@Fox2015], cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020], and shape and symmetry standardization of British Acheulean [@Hoggard2019].

@@

We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)). All drawings provide a two-dimensional view of vessel cross-sections, and metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced. In those instances where only one side of the cross-section image was available, these were mirrored and joined to create a closed outline for each vessel. Analyses were conducted in R software (R Core Team, 2019) using functions included in the Momocs package for quantifying and analyzing shapes [@Bonhomme2014]. Outlines were converted into a list of successive x, y pixel coordinates for EFA. We analyzed harmonic coefficients by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

@@

<!-- size analysis -->
We also conducted size analysis by examining the body diameter of the vessels to explore any differences in vessel form reflected by size and shape. Body diameter is a suitable metric measurement to represent size for ceramics since it is associated with volume as a utilitarian function, and could present differences in the situation of similar shape. The body diameter of the vessels from Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A). It is interesting that linear regression model (Figure \@ref(fig:size-shape-regression-plots): B) predicts that shape, captured by PC1 and PC2, changes with size across three phases, except PC1 in Chinese period. That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size; however, the vessel from Chinese period does not show this pattern. On the contrary, the negative relationship between PC2 and size suggests that the vessels with narrower neck and mouth tend to have a larger body diameter.

@@

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community. Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Shape changes are more pronounced during the Chinese contact period. In general, vessels become shorter in height over time, leading to an oval-shaped body. Moreover, size significantly varies with shape for vessels before and after the European presence, which presents that shorter vessels with narrower opening tend to have a larger body diameter. This correlation between size and shape can be also observed during the Chinese period without the consideration of height.

>>>>

Geometric morphometrics differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points that can be specified on a coordinate system. Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone tools or iron weapons [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017]. With univariate or multivariate statistics, the GMM applications can answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], and hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019]. A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent overall shape. Visualisation of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them.

Key questions in archaeological shape analysis often involve measuring shape standardization over time, or between geographical areas. Standardization is often investigated using multivariate analysis on shape variables computed from landmark data <!-- make sure it's clear that this is a common element in the studies in this para -->, or coefficients of variation on associated metric data, especially for lithic assemblages. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points. @Buchanan2018 studied metric data and GMM to identify a more uniform base-shape of Folsom points than Clovis points across the western US [@Buchanan2019]. Similarly, @Smith2016 suggested that the fluted point base in Alaska and northern Yukon is a risk management strategy to ensure the ease of replacement during long-distance travel <!-- not clear how this fits in this paragraph--> . Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014]. For ceramics, @Topi2017 found standardized vessel shapes from the Casas Grandes culture in northwest Mexico using coefficients of variation for the positions of semi-landmarks across shape groups. They suggest standardization might indicate the presence of specialized producers, reflecting social complexity. Another way to examine standardization is using morphological disparity to perform pairwise comparisons of correlations between shape groups [@Wills2001] <!-- this is vague --> . In this manner, @Selden2018ceramic2; @Selden2019 examined Caddo ceramics in northeast Texas and found a gradual increase in shape standardization over time that provides a basis for further discussion of craft specialization or group identity <!-- make it clear he did something different to you -->. Similar applications to bifaces in central Texas show greater diversity in morphology <!-- any interpretation? --> [@Selden2020]. Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances is an effective method to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].

Despite the wide use of landmark approaches in archaeology, a key limitation is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, outline approach, such as Elliptic Fourier Analysis (EFA), is a better mean for assessing morphological variation in whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. To be specific, EFA converts coordinates along a curve into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. The first systematic use of EFA in archaeology was @Gero1984's study of lithic flakes in Peru that suggests the changes in forms from a more angular to rounded shape over time. Later, @Saragusti2005 introduced several functions for the further calculation of specific shape attributes, including symmetry, roughness, and deformation, which demonstrates the potential for analysis of curves. To compare resharpening trajectories in European Middle Paleolithic stone tools, @Iovictua2009 presented a protocol from outline digitization to EFA and multivariate linear regression. More recent applications to stone artifacts seek to answer questions about typological classification of Late Woodland points [@Fox2015], cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020], and shape and symmetry standardization of the British Acheulean [@Hoggard2019].

@@

We used scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)). All drawings provide a two-dimensional view of vessel cross-sections, and metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced. In those instances where only one side of the cross-section image was available, or some sections were missing, we interpolated the curves and then mirrored and joined to create a closed outline for each vessel. Analyses were conducted in R software (R Core Team, 2019) using functions included in the Momocs package for quantifying and analyzing shapes [@Bonhomme2014]. Outlines were converted into a list of successive x, y pixel coordinates for EFA. We analyzed harmonic coefficients by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

@@

<!-- size analysis -->

We also conducted size analysis by examining the body diameter of the vessels to explore any differences in vessel form reflected by size and shape. We measured body diameter directly from each physical vessel in the collection, and we focus on this metric because it is available for more vessels than any other metric. The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A). Our linear regression models (Figure \@ref(fig:size-shape-regression-plots): B) show that shape, captured by PC1 and PC2, changes with size across the three phases, except for PC1 in the Chinese period. That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size. However, vessels from the Chinese period does not show this pattern. On the contrary, the negative relationship between PC2 and size suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.

@@

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community. Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Shape changes are more pronounced during the Chinese contact period. In general, vessels become shorter in height over time, leading to an oval-shaped body. Moreover, size significantly varies with shape for vessels before and after the European presence. Shorter vessels with a narrower opening tended to have a larger body diameter. This correlation between size and shape can be also observed during the Chinese period, but there is no relationship with height.
